Contracts for unions and intersections are hard for Nickel

### 2021-05-27 paper.tex
Some more sec 4

Type systems featuring intersection types are hard in a fundamental way,
from their discovery, they've been studied as a way to characterize
lambda-calculus terms normalizability; which means that typechecking a
term on a system with intersection types is undecidable,
similar to how the halting problem is undecidable\todo{reference}.

@@

As we intend to show on this paper, unions and intersections are not only
statically hard to check, but they are also dynamically hard;
\emph{they are fundamentally hard}.
In this section, we present intuitively the main challenges that arise
when attempting to develop a runtime check system for unions and intersections.

\subsubsection*{Union contracts are hard to break}

When programming with dynamic checks it is assumed that lack of
failure does not mean that the program works perfectly.
For instance, imagine having a function wrapped in a contract
that checks that it maps positive numbers to positive numbers, like
function \nickel{f} on Figure \ref{fig:pos-to-pos}.

Clearly, there is an error on this program, but if the first call
to \nickel{f} that gets executed is \nickel{f 10}, the error will not
be found at that point, and the first operand of the addition would be
evaluated to \nickel{3}.
If, later on, \nickel{f 5} gets executed, the contract would get violated,
since \nickel{f} gets called with a positive value, but returns a negative
one (\nickel{-2}).

@@

Having a witness of size 1 for every contract failure is a nice to have property
since it means, for example, that every time that a contract is violated,
the stack trace is enough to understand how it was violated, and to potentially
come up with a minimal example that breaks the contract.
However, when unions get introduced, it is lost.
Consider now the program presented in Figure \ref{fig:wrong-union-function},
this program (and in particular, \nickel{f}'s contract) still presents a problem,
since \nickel{fun x => x - 7} does not map positive numbers to positive numbers,
nor does it map positive numbers to non-positive numbers, so it won't be a valid
value for the union of those contracts.

@@

Still, the program presents enough function calls to find that issue
and violate the contract,
since \nickel{f 10} is a witness of \nickel{f} failing \nickel{Positive -> NonPositive},
and \nickel{f 5} is a witness of \nickel{f} failing \nickel{Positive -> Positive}.
But, what makes this case completely different from the one presented before, is that
both function calls are needed, and, what is more, there is not a way to
use the wrapped value \nickel{f} only once that would find a violation to its contract;
at least two are needed.

This seemingly simple consequence generates major issues when considered in real
life languages.
For example, remember the different mechanisms for code manipulation presented
in Section \ref{sec:optimizations}, inlining and commmon subexpression elimination,
both of these rely on the fact that substituting a pure function call by its body,
or joining equal pure expressions and sharing their result, is a semantically valid
transformation.
However, consider the examples presented at Figure \ref{fig:optimized-programs},
the first one is the inlined version of the second one, while the second
one has something like a CSE applied to it.

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Inlined]
let f = (fun x y => if y then x else "False") @ Number -> (Boolean -> Number) | (Boolean -> String) in
(f 1 true, f 1 false)
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Common Subexpression Eliminated]
let f = (fun x y => if y then x else "False") @ Number -> (Boolean -> Number) | (Boolean -> String) in
let g x = f 1 x in
(g true, g false)
\end{lstlisting}
\caption{Equivalent programs, with inlining or CSE applied}
\label{fig:optimized-programs}
\end{figure}
\todo{Improve example}

Unexpectedly, this two programs behave differently, the first one will return
\nickel{(1, "False")}, without failing.
While the second one will fail, since the single application of \nickel{f 1},
applied to both \nickel{true} and \nickel{false} will be able to find the error.

Summing it up, wrapping a value with a union of higer order contracts means that
that value now has a state, that may get modified silently at every use site.
Both reviewed papers suffer from this defect since it is the semantics
of unions that interact badly with code manipulation operations,
not a particular implementation.
This addition of global state complicates both the work of the language
implementor, as well as the language user, since it adds non trivial
side effects to your code.
\unsure{Maybe mention CSE is actualy commonly done by hand as well}

\subsubsection*{[WIP] Flat contracts are too permissive}

Stop for a second and think how you could implement an intersection
contract. The first thought you may have is "just apply
both contracts", as shown on the Implementation of Figure \ref{fig:inter-contracts}.
Of course, if it were that simple we wouldn't be writing this paper,
this idea starts to show problems when you first combine it with
higher order contracts, since it would mean not complying with
at least one of these contracts negatively would mean breaking the
whole thing, as shown on the second example on the same Figure.

@@

Still, not all hope is lost, we could pass some extra shared state
to each one of the sub contracts and let them only fail if the other
side of the contract has failed as well, as shown in Figure
\ref{fig:inter-contracts-2}, where the sared state is
represented by a label \nickel{l}.
This idea is still quite reasonable, for instance, when executing
\nickel{g 1}, from Figure \ref{fig:inter-contracts}, the first contract
that gets checked is the argument of \nickel{String -> String} over the function parameter, so
something like \nickel{1 @ String}.
Of course, this fails, but before raising a violation, this contract could
check if the other side of the intersection (\nickel{Number -> Number}) has
also tried to raise negative blame (on the parameter \nickel{1}), since
it hasn't, the contract would return normally, and would mark itself as failed.

@@

This idea, which is an oversimplification of the ideas
presented by Williams, Morris, and Wadler, seems to be spot on.
However, it presents a major problem, it's not compatible with flat contracts,
presented in Section \ref{sec:flat-contracts}.
Consider the example presented on Figure \ref{fig:inter-flat-contracts},
if the implementation presented previously were to be used, we may end up with
something like this \nickel{((fun x => x) @ (String -> String)[l]) @ C[l]}, where
\nickel{l} represents the shared state.
At this point, the intuitive thing to do is to check contract \nickel{C}, by checking
\nickel{((fun x => x) @ (String -> String)[l]) 0 == 0}.
This will fail, since applying \nickel{0} to a \nickel{String -> String} contract
fails negatively.

@@

But this should not have happened, since in the program of Figure \ref{fig:inter-flat-contracts}
\nickel{g} is never applied to \nickel{0}, in fact, it is never applied.
Checking if \nickel{C} has already failed, and marking \nickel{String -> String} as
failed negatively on \nickel{l} is not enough, the contract \nickel{String -> String}
should not have been used inside the evaluation of \nickel{C}, which
adds an extra layer of complexity, contracts should be guarded, and execution
depends on the shape of the surrounding evaluation context.

@@

Another difficult issue Keil and Thiemann were able to solve, is the
fact that flat contracts and intersection do not usually play nice
together.
They provide two different sets of rules for their calculus,
the first one is non deterministic, whose purpose is to provide
semantics that are simple to understand.
The second one is a deterministic set of rules, that are intended
to be simple to implement.
This second set needs to perform operations
on the evaluation context, to decide whether or not to evaluate
a given contract.
\unsure{Worth showing how the non deterministic version works? (See section B.1.1)}

>>>>

Type systems featuring intersection types are hard in a fundamental way,
from their discovery, they've been studied as a form to characterize
lambda-calculus terms normalizability; which means that typechecking a
term on a system with intersection types is undecidable,
similar to how the halting problem is undecidable\todo{reference}.

@@

As we intend to show on this paper, unions and intersections are not only
statically hard to check, but they are also dynamically hard:
\emph{they are fundamentally hard}.
In this section, we present intuitively the main challenges that arise
when attempting to develop a runtime check system for unions and intersections.

\subsection*{Union contracts are hard to break}

When programming with dynamic checks it is assumed that lack of
failure does not mean that the program works perfectly.

For instance, imagine having a function wrapped in a contract
that checks that it maps positive numbers to positive numbers,
and using it in different ways,
like the program on Figure \ref{fig:pos-to-pos}.
Clearly, there is an error on this program, but if the first call
to \nickel{f} that gets executed is \nickel{f 10}, the error will not
be found at that point, and the first operand of the addition would be
evaluated to \nickel{3}.
If, later on, \nickel{f 5} gets executed, the contract would get violated,
since \nickel{f} gets called with a positive value, but returns a negative
one (\nickel{-2}).

@@

Having a witness of size 1 for every contract failure is a really nice to have property
since it means, for example, that every time that a contract is violated,
the stack trace is enough to understand how it was violated, and to potentially
come up with a minimal example that breaks the contract.
However, when unions get introduced, this property is lost.

Consider now the program presented in Figure \ref{fig:wrong-union-function},
this program (and in particular, \nickel{f}'s contract) still presents a problem,
since \nickel{fun x => x - 7} does not map positive numbers to positive numbers,
nor does it map positive numbers to non-positive numbers, so it won't be a valid
value for the union of those contracts.

@@

As before, the program presents enough function calls to find that issue
and violate the contract,
since \nickel{f 10} is a witness of \nickel{f} failing \nickel{Positive -> NonPositive},
and \nickel{f 5} is a witness of \nickel{f} failing \nickel{Positive -> Positive}.
But, what makes this case completely different from the one presented before, is that
both function calls are needed, and, what is more, there is not a way to
use the wrapped value \nickel{f} only once that would find a violation to its contract;
at least two are needed.

This seemingly simple consequence generates major issues when considered in real
life languages.
For example, remember the different mechanisms for code manipulation presented
in Section \ref{sec:optimizations}, inlining and commmon subexpression elimination,
both of these rely on the fact that substituting a pure function call by its body,
or joining equal pure expressions and sharing their result, is a semantically valid
transformation.
However, consider the examples presented at Figure \ref{fig:optimized-programs},
the first one is an inlined version of the second one (where function
\nickel{g} has been inlined); while the second is similar to the first, but
the common subexpression \nickel{f 1} has been eliminated.

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Inlined]
let f = (fun x y => if y then x else "False") @ Number -> (Boolean -> Number) | (Boolean -> String) in
(f 1 true, f 1 false)
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Common Subexpression Eliminated]
let f = (fun x y => if y then x else "False") @ Number -> (Boolean -> Number) | (Boolean -> String) in
let g x = f 1 x in
(g true, g false)
\end{lstlisting}
\caption{Equivalent programs, with inlining or CSE applied}
\label{fig:optimized-programs}
\end{figure}

Unexpectedly, these two programs behave very differently, the first one will return
\nickel{(1, "False")}, without failing.
While the second one will fail, since the single application of \nickel{f 1},
applied to both \nickel{true} and \nickel{false} will be able to find the error
on the union contract (\nickel{(Boolean -> Number) | (Boolean -> String)}).

A valid argument against this point would be stating that further work should be done
on optimizers, so to not modify these tricky cases, even if this is true,
programmers also modify their code on a daily basis, and if moving from the first
code snippet at Figure \ref{fig:optimized-programs} to the second one changes
the behaviour of a program, we argue that union contracts not only go against the
implementation of languages, but also against their usability.

Summing it up, wrapping a value with a union of higer order contracts means that
that value now has a state, that may get modified silently at every use site.
Both reviewed papers suffer from this defect since it is the semantics
of unions that interact badly with code manipulation operations,
not a particular implementation.
This addition of global state complicates both the work of the language
implementor, as well as the language user, since it adds non trivial
side effects to your code.

\subsection*{Flat contracts do not intersect}

Stop for a second and think how you could implement an intersection
contract. The first thought you may have is "just apply
both contracts", as shown on the Implementation snippet
from Figure \ref{fig:inter-contracts}.
Of course, if it were that simple we wouldn't be writing this paper.
This idea starts to show problems when you first combine it with
higher order contracts, since it would mean not complying with
at least one of these contracts negatively would entail breaking the
whole thing, as shown on the second example on the same Figure,
that would fail.

@@

Not all hope is lost, we could pass some extra shared state
to each one of the sub contracts and let them only fail if the other
side of the contract has failed as well, as shown in Figure
\ref{fig:inter-contracts-2}, where the shared state is
represented by a label \nickel{l}.
This idea is still quite reasonable, for instance, when executing
\nickel{g 1}, from Figure \ref{fig:inter-contracts}, the first contract
that gets checked is the argument of \nickel{String -> String} over the function parameter, so
something like \nickel{1 @ String}.
Of course, this fails, but before raising a violation, this contract could
check if the other side of the intersection (\nickel{Number -> Number}) has
also tried to raise negative blame on the same parameter ()\nickel{1}), since
it hasn't, the contract would return normally, and would mark itself as failed.

@@

On the other hand, if you were to do something like
\nickel{g true}, that would fail also negatively the contract
\nickel{String -> String} and, when checking if \nickel{true}
is a correct parameter for \nickel{Number -> Number} it would check
if the other side of the intersection has failed, and since it
has, a violation would be raised and the whole program would
correctly fail.

This idea, which is an oversimplification of the ideas
presented by Williams, Morris, and Wadler, seems to be spot on.
However, it presents a major problem, it's not compatible with flat contracts,
as presented in Section \ref{sec:flat-contracts}.
Consider the example presented on Figure \ref{fig:inter-flat-contracts},
if the implementation discussed previously were to be used, we may end up with
something like this \nickel{((fun x => x) @ (String -> String)[l]) @ C[l]}, where
\nickel{l} represents the shared state.
At this point, the intuitive thing to do is to check contract \nickel{C}, by checking
if \nickel{((fun x => x) @ (String -> String)[l]) 0 == 0}.
This will fail, since applying \nickel{0} to a \nickel{String -> String} contract
fails negatively.

@@

But this should not have happened, since in the program of Figure \ref{fig:inter-flat-contracts}
\nickel{g} is never applied to \nickel{0}, in fact, it is never applied
to anything.
Checking if \nickel{C} has already failed, and marking \nickel{String -> String} as
failed negatively on \nickel{l} is not enough, the contract \nickel{String -> String}
should not have been used inside the evaluation of \nickel{C}, which
adds an extra layer of complexity: contracts should be guarded, and execution
depends on the shape of the surrounding evaluation context.

\subsection*{[WIP] Union and intersection are not a silver bullet}

Before wrapping up this section, we'd like to show that maybe, union and intersection
contracts are not exactly what we would want them to be.
As stated throughout this work, programmers coming from dynamic languages
are accustomed to perform safety checks (data validation) with
great amounts of flexibility, and the whole power of their
language of choice at their disposition.

For instance, a Python programmer may have no problem in checking
at runtime, that a given variable is an integer if a given flag is
not set, as shown in Figure \ref{fig:python-assertion},
while a Haskell programmer might close her eyes and leave the room
instantly.

\begin{figure}[h]
\begin{lstlisting}[language=python]
result = 42 if flag else "Disabled"
...
assert ((not flag) or isinstance(result, int))
\end{lstlisting}
\caption{Asserting at runtime the type of a variable}
\label{fig:python-assertion}
\end{figure}

These expectations are not really met by how unions and intersection
contracts are supposed to work.
For example, consider the Nickel code shown on Figure
\ref{fig:intersection-distrbution}, at first sight, this code should
complete execution succesfully, returning \nickel{(1, 1)},
however, under both the Keil and Thiemann, as well as the
Williams, Morris, and Wadler works, this program will fail, with a violation
of the contract imposed to \nickel{g}.

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let f = fun x y => x in
let g = f @ (Number -> Number -> Number) & (Number -> Boolean -> Number) in
let h = g 1 in 
(h 1, h true)
\end{lstlisting}
\caption{Intersection contracts don't distribute}
\label{fig:intersection-distrbution}
\end{figure}

The reason this happens is that
there is a fundamental characterization of how intersection
contracts behave: the choice of which branch to choose is made exactly
when the intersection is opened up, and not delayed as much as possible.
This means, that the contract presented on Figure \ref{fig:intersection-distrbution},
\nickel{(Number -> Number -> Number) & (Number -> Boolean -> Number)},
will not behave the same way that
\nickel{(Number -> (Number -> Number) & (Boolean -> Number)},
when these two expressions should be logically equivalent.

@@

Another difficult issue Keil and Thiemann were able to solve, is the
fact that flat contracts and intersection do not usually play nice
together.
They provide two different sets of rules for their calculus,
the first one is non deterministic, whose purpose is to provide
semantics that are simple to understand.
The second one is a deterministic set of rules, that are intended
to be simple to implement.
This second set needs to perform operations
on the evaluation context, to decide whether or not to evaluate
a given contract.
\unsure{Worth showing how the non deterministic version works? (See section B.1.1)
try with an example}
